---
title: "Package statss"
author: "Eric Ward"
date: "February 14, 2017"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tvvarss}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installation

```{r install, eval=TRUE, warning=FALSE, message=FALSE, results='hide'}
library(rstan)
library(devtools)
devtools::install_github("eric-ward/safs-timeseries/statss")
library(statss)
# for optimizing stan on your machine,
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

## Data

We'll use the same data as in the MARSS manual to illustrate an example of dynamic factor analysis (DFA)

```{r, eval=TRUE, warning=FALSE, message=FALSE, results='hide'}
library(MARSS)
# load the data (there are 3 datasets contained here)
data(lakeWAplankton)
# we want lakeWAplanktonTrans, which has been transformed
# so the 0s are replaced with NAs and the data z-scored
dat = lakeWAplanktonTrans
# use only the 10 years from 1980-1989
plankdat = dat[dat[,"Year"]>=1980 & dat[,"Year"]<1990,]
# create vector of phytoplankton group names
phytoplankton = c("Cryptomonas", "Diatoms", "Greens",
                   "Unicells", "Other.algae")
# get only the phytoplankton
dat.spp.1980 = plankdat[,phytoplankton]
```

## Estimation  

For starters, we can fit the trend model with 3 trends and no covariates.

```{r, warning=FALSE, message=FALSE, results='hide'}
# fit the three trend model with 3 covariates
y = t(dat.spp.1980)
model = fit_dfa(y = t(dat.spp.1980), num_trends=3, iter=1000)
```

### Varimax function

We can use the same varimax rotation as in MARSS to rotate the trends
```{r, warning=FALSE, message=FALSE, results='hide'}
rot = rotate_trends(model)
```

### Plot the trend estimates

```{r}
par(mfrow=c(2,2),mai=c(0.4,0.4,0.2,0.1))
for(i in 1:dim(rot$Z_rot_mean)[2]) {
  plot(rot$trends_mean[i,], type="l", main=paste("Trend:",i))
}
```

### Get the Z-estimates (loadings)

```{r}
print(rot$Z_rot_mean)
```

## Model selection

We can fit the models with 1-5 trends as follows. Note: all of these assume the observation variance is equal between time series (to change this, use the 'varIndx' argument).

```{r, warning=FALSE, message=FALSE, results='hide'}
mod_1 = fit_dfa(y = t(dat.spp.1980), num_trends=1, iter=500)
mod_2 = fit_dfa(y = t(dat.spp.1980), num_trends=2, iter=500)
mod_3 = fit_dfa(y = t(dat.spp.1980), num_trends=3, iter=500)
mod_4 = fit_dfa(y = t(dat.spp.1980), num_trends=4, iter=500)
mod_5 = fit_dfa(y = t(dat.spp.1980), num_trends=5, iter=500)
```

We can then use the `loo` package to extract the Leave One Out Information Criterion (LOOIC). We do this with the following, 

```{r}
library(loo)
loo(extract_log_lik(mod_1))$looic
```

## Covariates

Next as an example, we can illustrate how to fit a model with 4 covariates. These will be simulated with just noise. We can illustrate an example using 10 covariates with this dataset. 

Including covariates requires 2 matrices to be entered: the `covar` matrix to include the actual (raw covariate) values, and optionally the `covar_index` matrix, which specifies which parameter may be shared across time series. For example, we can force the effects of covariates 1-10 to be the same on species 1/2, the same on species 3/4, and unique for species 5. 
```{r, eval=FALSE, warning=FALSE, message=FALSE, results='hide'}
# fit the three trend model with 37 covariates
y = t(dat.spp.1980)
covar = matrix(runif(10*ncol(y)), nrow=10)
covar_index = matrix(1, nrow(y), nrow(covar))
# let's make species 1, 3 have the same effects, 2/4 have the same effects and 5 be different
covar_index[1,] = 1:10
covar_index[3,] = 1:10
covar_index[2,] = 11:20
covar_index[4,] = 11:20
covar_index[5,] = 21:30
fit_dfa(y = y, covar=covar, covar_index = covar_index, num_trends=3, iter=500)

```





